Musical elements in the discrete-time representation of sound

### 2017-08-09 article/acm/body.tex
finished final review until the end of spatialization

\subsection{Spatialization: localization and reverberation}\label{subsec:spac}
A musical note is always produced within the three dimensional physical space.
The consideration of this issue is the subject of the spatialization knowledge field and practice\footnote{By
spatialization one might both find: 1) the consideration of cues in the sound that derive from the environment,
including the localization of the listener and the sound source;
2) techniques to diffuse audio through various sources, such as loudspeakers, singers and traditional musical instruments, for musical purposes.
We focus in the first item although issues of the second item are also tackled.}.
A note is always spatialized
even though it is not one of its four basic properties in canonical musical theory
A note has a source which has a position at the ordinary three dimensional physical space.
This position is the spatial localization of the sound.
It is often (modeled as) a single point but can be a surface or a volume.
The reverberation in the environment in which a sound occurs is another one of the main topics of spatialization.
Both concepts, spatial localization and reverberation,
are widely valued by composers, audiophiles and the music industry~\cite{floEsp}.

\subsubsection{Spatial localization}
It is understood that the perception of sound localization occurs in our nervous system by three pieces of information: the delay of incoming sound between both ears, the difference of sound intensity at each ear and the filtering performed by the human body, specially its chest, head and ears~\cite{Roederer, hrtf, Heeger}.

@@

\noindent where $\zeta$ is the distance between ears $\zeta$, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in Interaural Time Difference:

@@

\noindent which, converted to amplitude, yields $IID_a=\frac{d}{d'}$. $IID_a$ can be used as a multiplicative constant to the right channel of a stereo sound signal: $\{t_i'\}_0^{\Lambda -1}=\{IID_a . t_i\}_0^{\Lambda -1}$, where $\{t_i'\}$ are samples of the wave incident in the left ear. It is possible to use IID together with ITD as a time advance for the right channel. It is a crucial clue to localization perception of bass sounds and percussive sonorities~\cite{Heeger}. 
With $\Lambda_{ITD}=\lfloor ITD . f_s \rfloor$:

@@

\noindent with $t_i$ as the right channel and $t_i'$ the left channel. If $\Lambda_{ITD} < 0 $, it is only needed to change $t_i$ by $t_i'$ and to use $\Lambda_{ITD}'= | \Lambda_{ITD} | $ and $IID_a'=1 / IID_a$.

@@

\noindent with $x,y$ as presented in Figure~\ref{fig:spac}. Henceforth, there are problems when $\theta$ falls within the so-called "cone of confusion": the same pair of ITD and IID results in a large number of points inside the cone. On those points the inference of the azimuthal angle depends especially on the attenuation filtering for high frequencies, since the head interferes much more in the treble than in bass waves~\cite{Heeger,hrtf}. Also relevant to the hearing of lateral sources is that low frequencies diffract and the wave arrives to the opposite ear with a delay of $\approx 0.7ms$.\cite{floEsp}

Figure~\ref{fig:spac} depicts the acoustic shadow of the cranium, an important phenomenon to perception of source azimuthal angle in the cone of confusion. The cone itself is not shown in Figure~\ref{fig:spac} because it is not exactly a cone and its precise dimensions were not encountered in the literature. Given the filtering and diffraction dependent on the sound spectrum, it is hard, if not impossible, to correctly draw the confusion cone. Even so, the cone of confusion can be understood as a cone with its apex placed in the middle of the head and growing out in the direction of each ear~\cite{hrtf}.

On the other hand, the complete localization, including height and distance of sound source, is given by the Head Related Transfer Function (HRTF)~\cite{hrtf}. There are well known open databases of HRTF, such as CIPIC, and it is possible to apply those transfer functions in a sonic signal by convolution (see Equation~\ref{eq:conv})~\cite{CIPIC}. Each human body has its filtering and there are techniques to generate HRTFs to be universally used~\cite{lazaSPA}.

\subsubsection{Spatialization}
Spatialization results from sound reflections and absorption by room/environment surface where the note is played. The sound propagates through the air with a speed of $\approx 343.2m/s$ and can be emitted from a source with any directionality pattern. When a sound pulse encounters a surface there are: 1) inversion of the propagation speed component normal to the surface;  2) energy absorption, especially in trebles. The sonic waves propagate until they reach inaudible levels (and further but then can often be neglected). As a sonic front reaches the human ear, it can be described as the original sound, with the last reflection point as the source, and the absorption filters of each surface it has reached. It is possible to simulate reverberations that are impossible in real systems. For example, it is possible to use asymmetric reflections with relation to the axis perpendicular to the surface, or to increase specific frequency bands (known as 'resonances'); neither of these characteristics are found in real systems.

There are reverberation models less related to each independent reflection, exploring valuable information to the auditory system. In fact, reverberation can be modeled with a set of two temporal and two spectral regions~\cite{JOSPhy}:
\begin{itemize}
   \item First period: 'first reflections' are more intense and scattered.
   \item Second period: 'late reverberation' is practically a dense succession of indistinct delays with exponential decay and statistical occurrences.
   \item First band: the bass has some resonance bandwidths relatively spaced.
   \item Second band: mid and treble have progressive decay and smooth statistical fluctuations.
\end{itemize}

Smith III states that usual concert rooms have total reverberation time of $\approx 1.9$ seconds, and that the period of first reflections is around $0.1s$. With these values, under the given conditions, there are perceived wave pulses which propagate for $652.08m$ ($83.79k$ samples in $f_s=44.1kHz$) before reaching the ear. In addition, sound reflections made after propagation for $34.32$ m ($4.41k$ samples in $f_s=44.1kHz$) have incidences less distinct by hearing. These first reflections are particularly important to spatial sensation. The first incidence is the direct sound, described by ITD and IID of Equations~\ref{eq:dti} and~\ref{eq:dii}. Assuming that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$, depending on the room dimensions, the separation between the first reflections is $8-90ms$ ($\approx 350-4000$ samples in $f_s=44.1kHz$). It is experimentally verifiable that the number of reflections increases with the square of $\approx k.n^2$. A discussion about the use of convolutions and filtering to favor implementation of these phenomena is provided in Section~\ref{subsec:mus2}, particularly in the paragraphs about reverberation.

>>>>

\subsection{Spatialization: localization and reverberation}\label{subsec:spac}
A musical note is always produced within the three dimensional physical space.
The consideration of this fact is the subject of the spatialization knowledge field and practice\footnote{By
spatialization one might find both:
1) the consideration of cues in sound that derive from the environment,
including the localization of the listener and the sound source;
2) techniques to produce sound through various sources, such as loudspeakers, singers and traditional musical instruments, for musical purposes.
We focus in the first item although issues of the second item are also tackled.}.
A note is always spatialized
even though it is not one of its four basic properties in canonical musical theory (duration, loudness, pitch and timbre).
A note has a source which has a position at the ordinary three dimensional physical space.
This position is the spatial localization of the sound.
It is often (modeled as) a single point but can be a surface or a volume.
The reverberation in the environment in which a sound occurs is another one of the main topics of spatialization.
Both concepts, spatial localization and reverberation,
are widely valued by composers, audiophiles and the music industry~\cite{floEsp}.

\subsubsection{Spatial localization}
It is understood that the perception of sound localization occurs in our nervous system mainly by three cues: the delay of incoming sound between both ears, the difference of sound intensity at each ear and the filtering performed by the human body, specially its chest, head and ears~\cite{Roederer, hrtf, Heeger}.

@@

\noindent where $\zeta$ is the distance between the ears, known to be $\zeta \approx 21.5cm$ in an adult human. Straightforward calculations result in Interaural Time Difference:

@@

$IID_a=\frac{d}{d'}$ can be used as a multiplicative constant to the right channel of a stereo sound signal: $\{t_i'\}_0^{\Lambda -1}=\{IID_a . t_i\}_0^{\Lambda -1}$, where $\{t_i'\}$ are samples of the wave incident in the left ear. It is possible to use IID together with ITD as a temporal advance for the right channel which is a crucial cue to localization perception of bass sounds and percussive sonorities~\cite{Heeger}. 
With $\Lambda_{ITD}=\lfloor ITD . f_s \rfloor$:

@@

\noindent where $t_i$ is the right channel and $t_i'$ is the left channel.
If $\Lambda_{ITD} < 0$, it is necessary to change $t_i$ by $t_i'$
and to use $\Lambda_{ITD}'= | \Lambda_{ITD} | $ and $IID_a'=1 / IID_a$.

@@

\noindent with $x,y$ as presented in Figure~\ref{fig:spac}.
Furthermore, there are problems when $\theta$ falls within the so-called "cone of confusion": the same pair of ITD and IID are related to a large number of points inside the cone~\footnote{The cone itself is not shown in Figure~\ref{fig:spac} because it is not exactly a cone and its precise dimensions were not found in the literature. Given the filtering and diffraction dependent on the sound spectrum, it is hard, if not impossible, to correctly draw the cone. Even so, the cone of confusion can be pictured as a cone with its apex placed in the middle of the head and growing out in the direction of the ear~\cite{hrtf}.}.
On those points the inference of the azimuthal angle depends especially on the filtering of the frequencies by the head and torso~\cite{Heeger,hrtf}.
Figure~\ref{fig:spac} depicts the acoustic shadow of the cranium, an important phenomenon to perception of source azimuthal angle in the cone of confusion. Also relevant to the hearing of lateral sources is that low frequencies diffract and the wave arrives to the opposite ear with a delay of $\approx 0.7ms$~\cite{floEsp}. 
The complete localization, including height and distance of sound source, is given by the Head Related Transfer Function (HRTF)~\cite{hrtf}. There are well known open databases of HRTF, such as CIPIC, and it is possible to apply those transfer functions in a sonic signal by convolution (see Equation~\ref{eq:conv})~\cite{CIPIC}. Each human body has its filtering and there are techniques to generate HRTFs to be universally used~\cite{lazaSPA}.

\subsubsection{Reverberation}
The reverberation results from sound reflections and absorption by the environment (e.g. a room) surface where a sound occurs.
The sound propagates through the air with a speed of $\approx 343.2m/s$ and can be emitted from a source with any directionality pattern. When a sound pulse encounters a surface there are: 1) inversion of the propagation speed component normal to the surface;  2) energy absorption, especially in high frequencies.
The sonic waves propagate until they reach inaudible levels (and further but then can often be neglected).
As a sonic front reaches the human ear, it can be described as the original sound, with the last reflection point as the source, and the absorption filters of each surface it has reached.
It is possible to simulate reverberations that are impossible in real systems. For example, it is possible to use asymmetric reflections with relation to the axis perpendicular to the surface, or to increase the intensity of the last reflections.

There are reverberation models less related to each independent reflection and that explores valuable cues to the auditory system. In fact, reverberation can be modeled with a set of two temporal and two spectral regions~\cite{JOSPhy}:
\begin{itemize}
   \item First period: 'first reflections' are more intense and scattered.
   \item Second period: 'late reverberation' is practically a dense succession of indistinct delays with exponential decay and statistical occurrences.
   \item First band: the bass has some resonance bandwidths relatively spaced.
   \item Second band: mid and treble have a progressive decay and smooth statistical fluctuations.
\end{itemize}

Smith III states that usual concert rooms have total reverberation time of $\approx 1.9$ seconds, and that the period of first reflections is around $0.1s$. With these values, under the given conditions, there are perceived wave pulses which propagate for $652.08m$ ($83.79k$ samples in $f_s=44.1kHz$) before reaching the ear. In addition, sound reflections made after propagation for $34.32$ m ($4.41k$ samples in $f_s=44.1kHz$) have incidences less distinct by hearing. These first reflections are particularly important to spatial sensation. The first incidence is the direct sound, described by ITD and IID of Equations~\ref{eq:dti} and~\ref{eq:dii}. Assuming that each one of the first reflections, before reaching the ear, will propagate at least $3-30m$, depending on the room dimensions, the separation between the first reflections is $8-90ms$ ($\approx 350-4000$ samples in $f_s=44.1kHz$).
Also, it is experimentally verifiable that the number of reflections increases with the square of time.
A discussion about the use of convolutions and filtering to favor implementation of these phenomena is provided in Section~\ref{subsec:mus2}, particularly in the paragraphs about reverberation.
